Think deep for reef resilience

### 2025-07-11 paper/paper_draft.qmd
qualify "range"
Standardise methods to past tense

The spatial extent of the GBR coral reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority
\[https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::great-barrier-reef-features-/explore?location=-19.221541%2C145.776418%2C4.94\].
In total this covers 3806 reef areas across the GBR marine park. Annually and spatially
averaged Degree Heating Week (DHW) data was used to represent future thermal stress at each
reef location.

Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great
Barrier Reef were obtained from the HighResCoralStress database [@dixon2022; @dixon].
Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology
[@liu2017; @skirving2020]. The SSTs are represented by a selection of five Global Climate
Models (GCMs) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6;
**citation**). These five GCMs were EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and
GDFL-CM4, and their inclusion were based on evaluation by Grose et al. [-@grose2023], which
found high correlation between observed temperature, rainfall, El Niño Southern Oscillation
and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding
mechanisms in these GCMs.
<!-- @Ben, the last sentence above is quite long and doesn't explain why the five GCMs
were selected. -->

Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming
caused by a doubling of carbon dioxide above its pre-industrial concentration
[@forster2021]. The four GCMs <!-- @Ben, says "four GCMs" here, but we used five? --> span a
range of ECS values within the IPCC assessed likely range, and GCM development families,
allowing for a range of possible climate futures to be explored, capturing the uncertainty
in future responses to CO2 increases (@fig-ecs; [@grose2023; @shepherd2018]). The following
analyses are performed using each GCM separately. SSP 2-4.5 was
<!-- I changed "is" to "was" - anything to do with the analysis making the decision to
standardise to past tense as currently there is a mix. --> used as a plausible emission
pathway under current emission trajectories, with similar atmospheric CO2 levels remaining
until 2050 [@hausfather2020; IPCC -@climate2023].

@@

We track the thermal tolerance of each functional group and size class as a truncated normal
distribution that is updated in each timestep following growth and recruitment. This
distribution determines that proportion of the population on a reef that is susceptible to
bleaching (i.e. has a thermal threshold below the experienced DHW value). Starting
distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021].

Tropical cyclones and impact of wave activity to reefs are not included in this
investigation. We scale coral bleaching impacts by a coefficient determined by the median
depth of each reef location, following Baird et al. [@baird2018], reducing the proportion of
coral colonies on a reef impacted by high heat stress as depth increases. Median depths for
each reef location have been extracted from satellite derived bathymetry data for the Great
Barrier Reef (GBRMPA [-@gbr10gb2020]) (see discussion for more details). Due to varying
degrees of alignment between bathymetry data and the defined outlines for GBR reefs, reefs
with less than 5% coverage of valid bathymetry data have been removed from this
investigation. Furthermore, reefs that were outside the GBRMPA management areas or not
associated with a bioregion were removed from the investigation.

@@

Given the uncertainties present, the proportion of reefs that display capacity to maintain
coral cover above a given $\beta_{0}$ threshold is examined. An indicative range of 10 to
20% is used (inclusive, with a step size of 1%). The range was informed by Perry et al.,
[-@perry2015] and Desbians et al., [-@desbiens] providing plausible indications of the
lower and upper bounds respectively. Robustness of the conclusions drawn to the uncertainty
in the precise value of $\beta_{0}$ for all reefs can then be assessed by using a range.

@@

Reefs in each timeseries cluster are compared to identify characteristics that differ
between low, medium and high coral cover clusters. These characteristics include reef median
depth, mean timeseries DHW, mean timeseries DHW – tolerance level, Pearson’s correlation
between total coral cover and inversed DHW timeseries, proportion of coral habitable area
occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and
total strength of larval connections between reefs. To analyse the DHW - tolerance levels of
reefs, heritable DHW – tolerance values were tracked over timeseries in a timestep \* taxa
\* reef \* scenario matrix. The mean timeseries was calculated over the first four scenarios
for each GCM (four from 4096 scenarios per GCM) due to computational restraints, and then
used to calculate an overall mean DHW – tolerance value for each reef under each GCM.

>>>>

The spatial extent of the GBR coral reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority
\[https://geohub-gbrmpa.hub.arcgis.com/datasets/GBRMPA::great-barrier-reef-features-/explore?location=-19.221541%2C145.776418%2C4.94\].
In total this covers 3806 reef areas across the GBR marine park. Annually and spatially
averaged Degree Heating Week (DHW) data were used to represent future thermal stress at each
reef location.

Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great
Barrier Reef were obtained from the HighResCoralStress database [@dixon2022; @dixon].
Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology
[@liu2017; @skirving2020]. DHWs were represented by a selection of five Global Climate
Models (GCMs) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6;
**citation**). These five GCMs were EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and
GDFL-CM4, and their inclusion were based on evaluation by Grose et al. [-@grose2023]. Grose et al. [-@grose2023] found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs.
<!-- @Ben, the last sentence above is quite long and doesn't explain why the five GCMs
were selected. -->

Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming
caused by a doubling of carbon dioxide above its pre-industrial concentration
[@forster2021]. The five GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; [@grose2023; @shepherd2018]). The following analyses were performed using each GCM separately. SSP 2-4.5 was
used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 [@hausfather2020; IPCC -@climate2023].

@@

In ADRIA we track the thermal tolerance of each functional group and size class as a truncated normal distribution that is updated in each timestep following growth and recruitment. This
distribution determines that proportion of the population on a reef that is susceptible to
bleaching (i.e. has a thermal threshold below the experienced DHW value). Starting
distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021].

Tropical cyclones and impact of wave activity to reefs were not included in this
investigation. We scale coral bleaching impacts by a coefficient determined by the median
depth of each reef location, following Baird et al. [@baird2018], reducing the proportion of
coral colonies on a reef impacted by high heat stress as depth increases. Median depths for
each reef location were extracted from satellite derived bathymetry data for the Great
Barrier Reef (GBRMPA [-@gbr10gb2020]) (see discussion for more details). Due to varying
degrees of alignment between bathymetry data and the defined outlines for GBR reefs, reefs
with less than 5% coverage of valid bathymetry data were removed from this
investigation. Furthermore, reefs that were outside the GBRMPA management areas or not
associated with a bioregion were removed from the investigation.

@@

Given the uncertainties present, the proportion of reefs that display capacity to maintain
coral cover above a given $\beta_{0}$ threshold is examined. An indicative range of 10 to
20% is used (inclusive, with a step size of 1%). The range was informed by Perry et al.,
[-@perry2015] and Desbians et al., [-@desbiens] providing plausible indications of the
lower and upper bounds respectively. Robustness of the conclusions drawn to the uncertainty
in the precise value of $\beta_{0}$ for all reefs can then be assessed by using the
indicative range.

@@

Reefs in each timeseries cluster were compared to identify characteristics that differ
between low, medium and high coral cover clusters. These characteristics include reef median
depth, mean timeseries DHW, mean timeseries DHW – tolerance level, Pearson’s correlation
between total coral cover and inversed DHW timeseries, proportion of coral habitable area
occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and
total strength of larval connections between reefs. To analyse the DHW - tolerance levels of
reefs, heritable DHW – tolerance values were tracked over timeseries in a timestep \* taxa
\* reef \* scenario matrix. The mean timeseries DHW - tolerance was calculated over the first four scenarios for each GCM (four from 4096 scenarios per GCM) due to computational restraints, and then used to calculate an overall mean DHW – tolerance value for each reef under each GCM.
